---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

In the repository, data is available as a .csv-file compressed into a .zip-file. If the zip-file has not been unpacked already, first unzip it. We then read the .csv file into a variable, which we shall call ```activity.data```.

```{r}
if (!file.exists('activity.csv')) 
    unzip("activity.zip")
activity.data <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?

```{r}
# Calculated value to be reported: first calculate total for each day, then calculate the mean
total_nr_of_steps_per_day <- aggregate(activity.data$steps, by=list(activity.data$date), FUN=sum)
mean_total_nr_steps_per_day <- mean(total_nr_of_steps_per_day$x, na.rm=TRUE)

# Setup reporting options
options(digits = 2, scipen = 6) # Prevent scientific notation,
```

The *mean* total number of steps taken per day is **`r mean_total_nr_steps_per_day`**.

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
